---
title: "niceR"
author: "Stefano Meschiari"
date: "June 21, 2016"
output:
  md_document:
    variant: markdown_github
---
`niceR` makes the R language a bit nicer by adding a few conveniences to the base language and a few new convenience functions and data structures. 

Install the package using [devtools](http://cran.r-
project.org/package=devtools):
```{r, eval=FALSE}
devtools::install_github('stefano-meschiari/nicer')
```
and import the library with `library`:
```{r, eval=FALSE}
library(nicer)
```
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(nicer)
```

License is MIT.

# NiceR features
The current version of `niceR` offers the following features:

- `+` works as a string concatenation operator, `%%` as a string formatting operator.
- a proper dictionary class with character keys, called `dict`.

## A proper dictionary class
The new class `dict` can be used to represent a heterogeneous dictionary with character keys. `dict`s have a few advantages over named lists:

- Every value is uniquely associated with a key (no holes).
- Keys are unique and cannot be repeated.
- Printing of dicts is more compact.
- Keys are never partially matched (a key named `test` will not match with `t` or `te`).
- A dict can have default values for non-existing keys.

### Creating a dictionary
The `dict` function can be used to create a new dictionary:
```{r}
d <- dict(color='blue', pattern='solid', width=3)
```

You can create a dictionary out of a list of keys and values using `make_dict`: 
```{r}
d <- make_dict(keys=c('color', 'pattern', 'width'),
               values=c('blue', 'solid', 3))
```

You can convert a named list to a dictionary using `as.dict`:
```{r}
d <- as.dict(list(color='blue', pattern='solid', width=3))
```

Printing looks nice:
```{r}
print(d)
```

### Accessing entries
You can access values using the familiar R syntax for lists:
```{r, eval=FALSE}
d <- dict(color='blue', pattern='solid', width=3)

d$color                   # blue
d[['pattern']]            # solid
d[c('color', 'pattern')]  # a sub-dictionary with keys color and pattern
```

You can get keys and values using the `keys` and `values` functions:
```{r, eval=FALSE}
keys(d)      # c('color', 'pattern', 'width')
values(d)    # list('blue', 'solid', 3)
```

You can get a list of "entries" (each element being a list with key and value). This is useful for iteration:
```{r, eval=FALSE}
for (entry in entries(d)) {
  cat(entry$key, ' = ', entry$value)
}
```

You can use the `mapkv` and `mapkv_dict` to map over keys and values. These functions are equivalent to the `map` family of functions in [purrr](https://github.com/hadley/purrr).

```{r, eval=FALSE}
# Returns a list
mapkv(d, function(k, v) paste(key, as.character(v)))
# Returns a dict with the same keys and squared values
d2 <- dict(a=1, b=2, c=3)
mapkv_dict(d, function(k, v) v^2)
```

### Utilities
A few utility functions inspired by the Underscore library:

- `invert(dict)` returns a dictionary where keys and values are swapped.
- `has(dict, key)` returns TRUE if `dict` contains `key`.
- `omit(dict, key1, key2, ...)` returns a new dictionary omitting all the specified keys.
- `extend(dict, dict1, ...)` copies all entries in `dict1` into `dict`, overriding any existing entries and returns a new dictionary.
- `defaults(dict, defaults)` fill in entries from `defaults` into `dict` for any keys missing in `dict`.

### Default dictionaries
You can create a dictionary with default values using `default_dict`. Any time a non-existing key is accessed, the default value is returned.
```{r}
salaries <- default_dict(employee_a = 50000,
                         employee_b = 100000,
                         default = 65000)
```

You can turn any dictionary into a default dictionary by setting its `default` attribute:

```{r}
attr(salaries, 'default') <- 70000
```

### Strict dictionaries
You can create a strict dictionary using `strict_dict`. Any time a non-existing key is accessed, an exception is thrown using `stop()`.
```{r}
# Associating each letter with a number
letters_to_numbers <- strict_dict(a=1, b=2, c=3, d=4) # etc.

# Accessing an existing key, that's OK!
print(letters_to_numbers$a)

# Accessing a non-letter will throw!
tryCatch(letters_to_numbers$notaletter, error=function(e) print(e))
```


## String concatenation using `+`
After you import `niceR`, you can concatenate strings with the regular `+` operator. The `+` operator has the same semantics as the `paste0` or `stringr::str_c` functions.

```{r}
a <- 'My string'
b <- 'another string'
print(a + ' and ' + b)

nums <- c(exp(1), pi)
print('My favorite number is ' + nums)
```

## String formatting using `%%`
The `%%` operator can be used in place of `sprintf`. The right operand should be a vector or list.

```{r}
print('New random number: %.2f' %% runif(1))

name <- 'Stefano'
last_name <- 'Meschiari'
print('Welcome, %s %s' %% c(name, last_name))
```

