---
title: "dictr: A new dictionary class for R"
author: "Stefano Meschiari"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dictr: A new dictionary class for R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, setup, include=FALSE}
library(dictr)
library(purrr)
```
`dictr` provides a new data structure specialized for representing dictionaries with string keys. 

Install the package using [devtools](https://cran.r-project.org/package=devtools):
```{r, eval=FALSE}
devtools::install_github('stefano-meschiari/dictr')
```
and import the library with `library`:
```{r, eval=FALSE}
library(dictr)
```


# `dict`: a dictionary class
In R, the typical way to store dictionaries is to use *named lists*:
```{r}
my_dict <- list(color='blue', pattern='solid', width=3)
my_dict$color 
```

Unfortunately, named lists have a host of issues when used to represent 
dictionaries: among others... 

* they can mix character and integer keys;
* they will do *partial matching*: by default (`my_dict$co` will also return `blue`, since it partially matches the key `color`);
* keys can be missing, non-unique, `NA`.

The new class `dict` can be used to explicitly model a dictionary with character keys. 
`dict`s have a few advantages over named lists:

- Every value is uniquely associated with a string key (no holes).
- Keys are unique and cannot be repeated.
- Keys are never partially matched (a key named `test` will not match with `t` 
  or `te`).
- Printing of dicts is more compact.
- A dict can have default values for non-existing keys.
- Dicts can contain NULL values.

Unlike other packages, `dict` uses regular lists under the hood.

This library also provides a rich library of useful functions for manipulating 
dictionary-like objects.

## Creating a dictionary
The `dict` function can be used to create a new dictionary:
```{r}
d <- dict(color='blue', pattern='solid', width=3)
```

You can create a dictionary out of a list of keys and values using `make_dict`: 
```{r}
d <- make_dict(keys=c('color', 'pattern', 'width'),
               values=c('blue', 'solid', 3))
```

You can convert a named list to a dictionary using `as_dict`:
```{r}
d <- as_dict(list(color='blue', pattern='solid', width=3))
```

Printing looks nice:
```{r}
print(d)
```

You can use a shorthand syntax that creates implicit keys based on the name of 
the variables:

```{r}
operating_system <- 'Amiga OS'
version <- '3.1'
cpu <- '68040'

machine <- dict(operating_system, version, cpu)

print(machine)
```

## Accessing entries

You can access values using the familiar R syntax for lists:

```{r, eval=FALSE}
d <- dict(color='blue', pattern='solid', width=3)

d$color                   # blue
d[['pattern']]            # solid
d[c('color', 'pattern')]  # a sub-dictionary with keys color and pattern
d['color', 'pattern']     # a sub-dictionary with keys color and pattern
```

You can get keys and values using the `keys` and `values` functions:
```{r, eval=FALSE}
keys(d)      # c('color', 'pattern', 'width')
values(d)    # list('blue', 'solid', 3)
```

You can get a list of "entries" (each element being a tuple with `key` and `value`) 
using the `entries` and `entry` functions. This is useful for iteration:
```{r, eval=FALSE}
for (entry in entries(d)) {
  cat(entry$key, ' = ', entry$value)
}
```

or to conveniently using apply-style functions:
```{r}
d <- dict(a=1, b=2, c=3)

d %>%
  entries() %>%
  map(~ entry(.$key, .$value^2)) %>%
  as_dict()
```

For convenience, you can use the `map_dict` function to map over keys and values, and the `keep_dict` and `discard_dict` to filter entries. These functions are specializations of the `map`, `keep` and `discard` family of functions in [purrr](https://purrr.tidyverse.org) that
take a dictionary and return a dictionary.

```{r, eval=FALSE}
# Returns a dict with the same keys and squared values
d <- dict(a=1, b=2, c=3)
map_dict(d, function(k, v) v^2)
```

## Setting entries

Entries can be set just like regular lists:

```r
d <- dict(first='Harrison', last='Solo')  # first=Harrison, last=Solo

d$last <- 'Ford'                          # first=Harrison, last=Ford
d[['first']] <- 'Leia'                    # first=Leia, last=Ford
d['last', 'title'] <- c('Organa', 'Princess') # first=Leia, last=Organa, title=Princess
```

Setting an entry to `NULL` does _not_ delete the entry, but instead sets the entry to `NULL`. To delete one or more entries, use the `omit` function:

```r
d <- dict(a=1, c=3)
d$b <- NULL         # a=1, b=NULL, c=3
d <- omit(a, 'a')   # b=NULL, c=3
```

## Utility functions
A few utility functions inspired by the Underscore library:

- `invert(dict)` returns a dictionary where keys and values are swapped.
- `has(dict, key)` returns TRUE if `dict` contains `key`.
- `omit(dict, key1, key2, ...)` returns a new dictionary omitting all the specified keys.
- `extend(dict, dict1, ...)` copies all entries in `dict1` into `dict`, overriding any existing entries and returns a new dictionary.
- `defaults(dict, defaults)` fill in entries from `defaults` into `dict` for any keys missing in `dict`.

The following functions specialize functions from the `purrr` library to dictionary objects:

- `map_dict(dict, fun, ...)` calls a function on each (key, value) pair and builds a dictionary from the transformed input.
- `keep_dict(dict, fun)` and `discard_dict(dict, fun)` keep or discard entries based on the function or predicate.
- `compact_dict(dict)` removes any entries with NULL values.

## Default dictionaries
You can create a dictionary with default values using `default_dict`. Any time a non-existing key is accessed, the default value is returned.
```{r}
salaries <- default_dict(employee_a = 50000,
                         employee_b = 100000,
                         default = 65000)
```

You can provide a default value for an existing dictionary by setting its `default` attribute:

```{r}
attr(salaries, 'default') <- 70000
```

## Strict dictionaries
You can create a strict dictionary using `strict_dict`. Any time a non-existing key is accessed, an exception is thrown using `stop()`.
```{r}
# Associating each letter with a number
letters_to_numbers <- strict_dict(a=1, b=2, c=3, d=4) # etc.

# Accessing an existing key, that's OK!
print(letters_to_numbers$a)

# Accessing a non-letter will throw!
tryCatch(letters_to_numbers$notaletter, error=function(e) print(e))
```

## Immutable collections
You can turn any collection (lists, vectors, and `dicts`) into an immutable collection using the `immutable` function. Such a collection cannot be modified using the `[`, `[[` and `$` operators; otherwise, it will behave the same as the original collection. The `immutable_dict` function creates an immutable dictionary.

```{r, error=TRUE}
const_letters <- immutable(letters)

# Will throw an error
const_letters[1] <- '1' 
```

```{r, error=TRUE}
physical_constants <- immutable_dict(
  astronomical_unit = 1.5e11,
  speed_of_light = 3e8,
  gravitational_constant = 6.7e-11
)

# Will throw an error
physical_constants$speed_of_light = 1
```

